---
title: "Project in Spatial Econometrics"
format: html
editor: visual
authors: "Anaïs BOUGHANEM, Théo DRUILHE and Sigurd SAUE"
---

We download the necessary packages.

```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(sf)
library(WDI)
library(tidyr) 
library(ggplot2)

```

# 1 Data

## 1.1 Migration flow data

We first load and analyze the data.

```{r}
# Load the migration data
load("mig_data.RData")

# Inspect the structure of the data
str(mig_data)
```

```{r}
# Filter for the period 2020–2024 and select the Pseudo-Bayesian flow estimates

mig_rate_2020_2024 <- mig_data %>%
  filter(period == "2020-2024") %>%
  select(origin, dest, mig_rate)

# Check the first few rows of the result
head(mig_rate_2020_2024)
```

### Migration Rate Method

Dennett (2016) proposed a method to estimate bilateral migration flows using a **single stock table** and an estimate of **total global migration**. This approach allows us to generate directional flows even when direct flow data is unavailable.

#### Step-by-Step Method

Let:

-   $s_{gh}^t$: Number of migrants **from origin** $g$ living in **destination** $h$ at time $t$
-   $M^t$: Estimated **total global migration** for the period
-   $r_{gh}^t$: Estimated **flow rate** from $g$ to $h$ at time $t$
-   $F_{g \rightarrow h}^t$: Estimated **flow** from $g$ to $h$

Then:

1.  **Compute global migration rate**: $$
    r_{gh}^t = \frac{s_{gh}^t}{\sum_{g \neq h} s_{gh}^t}
    $$ The denominator is the sum of all off-diagonal stock values — i.e., total foreign-born population globally.

2.  **Estimate total migration flows (M)**: Since $M^t$ is not directly known, it's approximated by: $$
    M^t = \sum_h \left| \sum_g s_{gh}^{t+1} - \sum_g s_{gh}^{t} \right|
    $$ This is the **sum of absolute net migration changes** per destination (column totals).

3.  **Estimate bilateral flow**: $$
    F_{g \rightarrow h}^t = M^t \times r_{gh}^t
    $$

#### Interpretation

-   The stock table provides the **relative distribution** of migrants.
-   The sum of absolute net changes provides a **scaling factor** for total migration volume.
-   The resulting flows maintain **global consistency** while reflecting the stock structure.

### Description of the Migration Rate Flow Map (2020–2024)

The map displays the largest bilateral migration flows for the period 2020–2024, estimated using the Migration Rate method. In this visualization, arrows represent the direction and relative magnitude of flows between countries, while vertical bars beside each country illustrate the total volume of outgoing (left bar) and incoming (right bar) flows.

This method emphasizes migration intensity, meaning flows are normalized by the size of the origin population. As such, it reveals not only where large numbers of people are moving, but also which countries are experiencing high migration rates relative to their population size.

Several prominent migration corridors are visible:

-   From Latin America to North America, especially strong flows from Mexico and Central America to the United States.
-   From South Asia and Southeast Asia to the Gulf countries or to the United States and Canada.
-   From Eastern Europe to Western Europe, especially toward Germany and the UK.

Major countries of origin with significant outward migration include India, Mexico, Syria and the Philippines. On the receiving side, major destinations include the United States, Germany, the United Kingdom, France, Australia, and the Gulf states.

The map highlights both the volume and directionality of global migration, showing that migration flows are not evenly distributed but rather concentrated in a few dominant corridors.

### Comparison of the Migration Rate Method with Other Estimation Methods

To evaluate the reliability and interpretability of the Migration Rate method, we compared it with five alternative estimation methods: Pseudo-Bayesian, Drop Negative, Reverse Negative, Minimization Open, and Minimization Close. The visualizations of each method reveal key similarities and differences in how major global flows are represented.

#### Similarities

Across all six maps, the same global migration corridors appear consistently, regardless of the estimation technique.

These consistent patterns suggest that the main global migration trends are robust across different estimation methods.

#### Notable Differences Between Methods

**Pseudo-Bayesian**

-   Flows seem globally bigger than on the other maps.

**Drop Negative and Reverse Negative**

-   These methods result in sparser maps with fewer visible corridors.
-   Strong flows like Mexico to the US still appear, but smaller flows are often eliminated.

**Minimization Open and Close**

-   Very similar to the Migration Rate method.

**Migration Rate**

-   The Russia to Ukraine arrow is reversed compared to the other methods, indicating a different interpretation of the flow direction.

## 1.2 Contours data

We load the contours data and inspect it.

```{r}
# Import the world contours GeoJSON file
world <- st_read("CCMN_GeoDATA.geojson")

# Transform the Coordinate Reference System to ESRI:54030 (World Robinson projection)
world_proj <- st_transform(world, crs = 'ESRI:54030')
```

## 1.3 Explanatory variables at country level

```{r}
load("covariates.RData")
```

## Additional Explanatory Variable: Unemployment Rate

To complement the existing set of explanatory variables, we propose adding the **unemployment rate** as a potential determinant of international migration. A high unemployment rate in a country can push individuals to emigrate in search of better opportunities, while low unemployment in destination countries may attract migrants seeking employment.

This variable is used in several studies, including Hatton & Williamson (2002) and the OECD International Migration Outlook reports, as a key economic factor in explaining migration trends.

The data was retrieved from the **World Bank Open Data** platform, using indicator code `SL.UEM.TOTL.ZS`, and merged into the `my_covariates` dataset for the years corresponding to each migration period (e.g., 1992 for 1990–1995).

```{r}
# Download unemployment rate from World Bank (indicator code: SL.UEM.TOTL.ZS)
# We'll select data for years matching the reference points (1992, 1997, 2007, 2012, 2017, 2022)
unemp_data <- WDI(indicator = "SL.UEM.TOTL.ZS", start = 1992, end = 2022, extra = TRUE) %>%
  filter(year %in% c(1992, 1997, 2002, 2007, 2012, 2017, 2022)) %>%
  select(iso3c, year, unemployment_rate = SL.UEM.TOTL.ZS)
```

```{r}
# Merge unemployment data into the covariates
unemp_data <- unemp_data %>%
  mutate(year = as.character(year))
my_covariates <- my_covariates %>%
  mutate(period = recode(period,
                         "1990-1995" = "1992",
                         "1995-2000" = "1997",
                         "2000-2005" = "2002",
                         "2005-2010" = "2007",
                         "2010-2015" = "2012",
                         "2015-2020" = "2017",
                         "2020-2024" = "2022"))
my_covariates <- my_covariates %>%
  left_join(unemp_data, by = c("ISO3"="iso3c", "period"="year"))
```

## 2 Migration Flows and Dependent Variables

In this section, we analyze both **immigration** and **emigration** flows, compute per capita migration rates, and create a net migration variable.

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)

# 1. Filtrer les données migration sur 2020–2024
mig_data_filtered <- mig_data %>%
  filter(period == "2020-2024")

# 2. Agréger les flux entrants et sortants
inflows <- mig_data_filtered %>%
  group_by(dest) %>%
  summarise(inflow = sum(reverse_neg, na.rm = TRUE))

outflows <- mig_data_filtered %>%
  group_by(origin) %>%
  summarise(outflow = sum(reverse_neg, na.rm = TRUE))

# 3. Garder uniquement la population pour 2022 (liée à 2020–2024)
pop_2022 <- my_covariates %>%
  filter(period == "2022") %>%
  select(ISO3, population)

# 4. Join clean
table_rates <- pop_2022 %>%
  left_join(inflows, by = c("ISO3" = "dest")) %>%
  left_join(outflows, by = c("ISO3" = "origin")) %>%
  mutate(
    inflow = replace_na(inflow, 0),
    outflow = replace_na(outflow, 0),
    immigrates = inflow / population,
    emigrates = outflow / population,
    net = immigrates - emigrates
  )

```

```{r}
# Top 10 pays en termes absolus (pas par habitant)
top10_outflow_abs <- table_rates %>%
  arrange(desc(outflow)) %>%
  slice_head(n = 10)

top10_inflow_abs <- table_rates %>%
  arrange(desc(inflow)) %>%
  slice_head(n = 10)

# Les deux graphes côte à côte
library(ggplot2)
library(patchwork)  # pour coller les graphes

p1 <- ggplot(top10_outflow_abs, aes(x = reorder(ISO3, outflow), y = outflow)) +
  geom_point(size = 3) +
  coord_flip() +
  labs(title = "Top Outflows", x = "", y = "") +
  theme_minimal()

p2 <- ggplot(top10_inflow_abs, aes(x = reorder(ISO3, inflow), y = inflow)) +
  geom_point(size = 3) +
  coord_flip() +
  labs(title = "Top Inflows", x = "", y = "") +
  theme_minimal()

# Afficher les deux plots côte à côte
p1 + p2

```

```{r}
# Ajouter les noms de pays
table_rates <- table_rates %>%
  left_join(world %>% st_drop_geometry() %>% select(ISO3, VISUALIZATION_NAME), by = "ISO3")

# Top 10 émigration/immigration par habitant
top10_emigrates <- table_rates %>%
  arrange(desc(emigrates)) %>%
  slice_head(n = 10)

top10_immigrates <- table_rates %>%
  arrange(desc(immigrates)) %>%
  slice_head(n = 10)

```

```{r}
# Points avec noms complets
p3 <- ggplot(top10_emigrates, aes(x = emigrates, y = reorder(VISUALIZATION_NAME, emigrates))) +
  geom_point(shape = 1, size = 3) +
  labs(title = "Top Outflows (per capita)", x = "", y = "") +
  theme_minimal()

p4 <- ggplot(top10_immigrates, aes(x = immigrates, y = reorder(VISUALIZATION_NAME, immigrates))) +
  geom_point(shape = 1, size = 3) +
  labs(title = "Top Inflows (per capita)", x = "", y = "") +
  theme_minimal()

# Affichage côte à côte
p3 + p4

```

```{r}
## 3 Correlation with Explanatory Variables

### Merge with explanatory variables (2022 only)

# Sélection des covariables explicatives pour 2022
cov_2022 <- my_covariates %>%
  filter(period == "2022") %>%
  select(
    ISO3, lifeexp, politicalstability, vulnerability, natural_disaster,
    conflictpercapita, t2m_diff, dry, prec_5days, ghwr_35, hwf_upp, unemployment_rate
  )

# Merge avec table_rates
table_corr <- table_rates %>%
  left_join(cov_2022, by = "ISO3")

### Correlation matrix

library(corrr)

# Sélection des colonnes numériques
numeric_vars <- table_corr %>%
  select(emigrates, immigrates, net,
         lifeexp, politicalstability, vulnerability, natural_disaster,
         conflictpercapita, t2m_diff, dry, prec_5days, ghwr_35, hwf_upp, unemployment_rate)

# Calcul des corrélations avec gestion des NA
cor_matrix <- correlate(numeric_vars, use = "pairwise.complete.obs")

# Visualisation
cor_matrix %>%
  stretch() %>%
  filter(x %in% c("emigrates", "immigrates")) %>%
  ggplot(aes(x = reorder(y, r), y = r, fill = x)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(title = "Correlation with Migration Rates",
       x = "Explanatory Variable", y = "Pearson correlation") +
  theme_minimal()


```

-   **Positive Correlates of Immigration:**
    -   **Life expectancy** and **political stability** are positively correlated with immigration. These countries tend to attract migrants due to better living conditions and governance.
    -   **Net migration** is, by construction, highly correlated with immigration.
    -   **Climatic variables** such as **temperature anomalies** and **precipitation extremes** show moderate positive relationships.
-   **Positive Correlates of Emigration:**
    -   **Conflict per capita** and **natural disasters** are positively associated with emigration, reflecting common push factors.
    -   **Drought conditions** and **heatwaves** (e.g., `dry`, `hwf_upp`) also align with climate-induced migration pressures.
-   **Inverse Relationships Across Migration Types:**
    -   **Political stability**: *positive* for immigration, *negative* for emigration.
    -   **Life expectancy**: *positive* for immigration, *negative* or weakly negative for emigration.
    -   **Vulnerability index**: tends to increase emigration and reduce immigration.

```{r}
# Corriger les géométries invalides et combiner les pays
world_clean <- world %>%
  mutate(geometry = st_make_valid(geometry)) %>%
  group_by(ISO3, VISUALIZATION_NAME) %>%
  summarise(geometry = st_union(geometry), .groups = "drop")

# Merge avec population + inflow/outflow
world_final <- world_clean %>%
  left_join(table_rates %>% select(ISO3, population, inflow, outflow, immigrates, emigrates, net), by = "ISO3")

```

```{r}
ggplot(world_final) +
  geom_sf(aes(fill = emigrates), color = "white", size = 0.1) +
  scale_fill_viridis_c(option = "inferno", name = "Taux d’émigration", trans = "sqrt") +
  theme_minimal() +
  labs(title = "Taux d’émigration par pays (2020–2024)")

```

```{r}
ggplot(world_final) +
  geom_sf(aes(fill = immigrates), color = "white", size = 0.1) +
  scale_fill_viridis_c(option = "plasma", name = "Taux d’immigration", trans = "sqrt") +
  theme_minimal() +
  labs(title = "Taux d’immigration par pays (2020–2024)")

```

```{r}
library(ggplot2)
library(RColorBrewer)

ggplot(world_final) +
  geom_sf(aes(fill = net), color = "white", size = 0.1) +
  scale_fill_gradient2(
    low = "red", mid = "black", high = "blue", midpoint = 0.2,
    name = "Net migration rate"
  ) +
  theme_void() +
  labs(
    title = "Net Migration Rate (2020–2024)",
    caption = "Blue = net inflow, Red = net outflow"
  )

```

```{r}
climate_2022 <- my_covariates %>%
  filter(period == "2022") %>%
  select(ISO3, t2m_diff)

world_final <- world_final %>%
  left_join(climate_2022, by = "ISO3")

ggplot(world_final) +
  geom_sf(aes(fill = t2m_diff), color = "white", size = 0.1) +
  scale_fill_viridis_c(option = "C", name = "Δ Température (°C)") +
  theme_void() +
  labs(
    title = "Variation de température (2020–2024)",
    caption = "Source: Données climatiques CCMN"
  )

```

```{r}
library(spdep)
library(sf)
library(ggplot2)

# 1. Préparer les centroids (coordonnées des centres de pays)
coords <- st_centroid(world_final$geometry)

# 2. Convertir les coordonnées en matrice
coords_mat <- st_coordinates(coords)

# 3. Créer les voisins contigus (Queen)
nb_contig <- poly2nb(world_final)

# 4. Créer les 4 plus proches voisins
nb_knn <- knn2nb(knearneigh(coords_mat, k = 4))

# 5. Combiner les deux
nb_combined <- union.nb(nb_contig, nb_knn)

# 6. Visualiser la carte avec les liens de voisinage
plot(st_geometry(world_final), col = "lightgray", main = "voisins")
plot(nb_combined, coords = coords_mat, add = TRUE, col = "black", lwd = 0.7)

```
### Moran scatterplot

```{r}
library(ggplot2)
library(dplyr)
library(spdep)

world_net <- world_final %>% 
  filter(!is.na(net))

# 2. Create spatial weights using only countries with data
coords_net <- st_coordinates(st_centroid(world_net$geometry))
nb_net <- poly2nb(world_net)
nb_net_combined <- union.nb(nb_net, knn2nb(knearneigh(coords_net, k = 4)))
listw_net <- nb2listw(nb_net_combined, style = "W", zero.policy = TRUE)
# 1. Prepare lagged variable and mean values
world_net$lag_net <- lag.listw(listw_net, world_net$net)
mean_net <- mean(world_net$net, na.rm = TRUE)
mean_lag_net <- mean(world_net$lag_net, na.rm = TRUE)

# 2. Quadrant classification
world_net$quad <- case_when(
  world_net$net >= mean_net & world_net$lag_net >= mean_lag_net ~ "High-High",
  world_net$net <  mean_net & world_net$lag_net <  mean_lag_net ~ "Low-Low",
  world_net$net >= mean_net & world_net$lag_net <  mean_lag_net ~ "High-Low",
  world_net$net <  mean_net & world_net$lag_net >= mean_lag_net ~ "Low-High"
)

# 3. Assign colors to quadrants
quad_colors <- c(
  "High-High" = "red",
  "Low-Low" = "purple",
  "High-Low" = "green",
  "Low-High" = "cyan"
)

# 4. Create the custom Moran scatter plot
ggplot(world_net, aes(x = net, y = lag_net, color = quad)) +
  geom_hline(yintercept = mean_lag_net, linetype = "dashed") +
  geom_vline(xintercept = mean_net, linetype = "dashed") +
  geom_point(size = 2, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  scale_color_manual(values = quad_colors, name = "Cluster") +
  theme_minimal() +
  labs(
    x = "Net Migration Rate",
    y = "Spatial Lag of Net Migration Rate",
    title = "Moran Scatter Plot: Net Migration Rate"
  )
```

```{r}
# 1. Reuse spatial weights already computed: listw_net
# 2. Local Moran's I on filtered data
local_moran_net <- localmoran(world_net$net, listw_net, zero.policy = TRUE)

# 3. Add results to the filtered spatial dataframe
world_net$local_I <- local_moran_net[, 1]         # Local Moran's I value
world_net$local_p <- local_moran_net[, 5]         # p-value
```

```{r}
library(dplyr)

# 1. Compute spatial lag
world_net$lag_net <- lag.listw(listw_net, world_net$net)

# 2. Define means
mean_net <- mean(world_net$net, na.rm = TRUE)
mean_lag_net <- mean(world_net$lag_net, na.rm = TRUE)

# 3. Compute Local Moran's I
local_net <- localmoran(world_net$net, listw_net, zero.policy = TRUE)
world_net$local_I <- local_net[, 1]
world_net$local_p <- local_net[, 5]

# 4. Define LISA quadrant
world_net$quad <- case_when(
  world_net$net >= mean_net & world_net$lag_net >= mean_lag_net ~ "HH",
  world_net$net <  mean_net & world_net$lag_net <  mean_lag_net ~ "LL",
  world_net$net >= mean_net & world_net$lag_net <  mean_lag_net ~ "HL",
  world_net$net <  mean_net & world_net$lag_net >= mean_lag_net ~ "LH"
)

# 5. Keep only significant clusters (e.g., p < 0.05)
world_net$quad_sig <- ifelse(world_net$local_p < 0.05, world_net$quad, NA)

# 6. Plot: significant LISA clusters
ggplot(world_net) +
  geom_sf(aes(fill = quad_sig), color = "white", size = 0.1) +
scale_fill_manual(
  values = c("HH" = "red", "LL" = "blue", "HL" = "green", "LH" = "cyan"),
  na.value = "grey90",  # <- this line fixes the error
  name = "LISA Cluster"
) +
  theme_minimal() +
  labs(
    title = "LISA Cluster Map: Net Migration Rate",
    subtitle = "Significant Local Spatial Autocorrelation (p < 0.05)",
    caption = "Red = High-High, Blue = Low-Low, Green = High-Low, Cyan = Low-High"
  )
```
### Moran test

```{r}
# Global Moran’s I test for net migration rate
moran_test_net <- moran.test(world_net$net, listw_net, zero.policy = TRUE)

# Print the test results
moran_test_net
```
#### Significance Test of Spatial Autocorrelation

We performed a Moran's I test to assess whether the spatial correlation in **net migration rates** (2020–2024) is statistically significant.

The test result is:

- **Moran’s I**: 0.165
- **Expected value** under null: –0.0047
- **Standard deviate (z-score)**: 4.22
- **p-value**: < 0.0001

These results indicate **significant positive spatial autocorrelation** in the net migration rate. That is, countries with high or low net migration rates are geographically clustered, justifying the use of spatial econometric models in further analysis.

### OLS Regression

```{r}
# Construire un dataset complet uniquement avec les colonnes disponibles
ols_data <- world_final %>%
  select(ISO3, VISUALIZATION_NAME, immigrates, emigrates, t2m_diff, population, inflow, outflow, net) %>%
  na.omit()

# Régression OLS — émigration
ols_emigrates <- lm(emigrates ~ t2m_diff + population + inflow + outflow + net, data = ols_data)

# Régression OLS — immigration
ols_immigrates <- lm(immigrates ~ t2m_diff + population + inflow + outflow + net, data = ols_data)

# Résumés des modèles
summary(ols_emigrates)
summary(ols_immigrates)
```
### Interpretation of the OLS results



#### Emigration Model (`emigrates`)

The model explains about **41%** of the variance in emigration rates across countries (**Adjusted R² = 0.40**), and the F-statistic confirms that the overall regression is **highly significant** (*p* < 0.001).

**Key findings:**
- **Temperature anomaly (`t2m_diff`)** has a **positive and significant** effect on emigration. A one-unit increase is associated with an increase of **0.0052** in the emigration rate.
- **Population** has a **negative effect**: more populous countries tend to show lower emigration rates per capita.
- **Outflow** is **positively associated** with emigration, as expected.
- **Net migration** has a **strong negative effect**, meaning countries with higher net inflow tend to have **lower** emigration rates.

#### Immigration Model (`immigrates`)

The immigration model performs better, explaining about **66%** of the variance in immigration (**Adjusted R² = 0.66**), with all coefficients highly significant.

**Key results:**
- **Temperature anomaly (`t2m_diff`)** is again a significant predictor of immigration.
- **Population** has a **negative** effect, consistent with per capita normalization.
- **Outflow** is **positively associated**, which may reflect circular or return migration patterns.
- **Net migration** has a **very strong positive effect** (**+0.743**, *p* < 0.001), indicating that countries with high net migration are also receiving proportionally more immigrants.

